Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism

### 2013-12-22 att-icfp14/att-icfp14.tex
Pasted ESOP paper into ICFP14

\documentclass{llncs}
\usepackage{amsmath}
\usepackage{llncsdoc}
\usepackage{amssymb} 
%\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
%\usepackage{cite}
%\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...

@@

\begin{document}
\mainmatter  % start of an individual contribution

% first the title is needed
\title{Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism}

% a short form should be given in case it is too long for the running head
\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Cyrus Omar%
\and Jonathan Aldrich}
%
\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Carnegie Mellon University, Pittsburgh, PA 15213, USA\\
\texttt{\{comar,aldrich\}@cs.cmu.edu}
}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

\toctitle{Lecture Notes in Computer Science}
\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
Researchers and domain experts often propose new language primitives as extensions to the semantics of an existing language. 
But today's statically-typed languages are {monolithic}: they do not expose language-internal mechanisms for implementing the static and dynamic semantics of new primitive types and their associated operators directly, so these experts must instead create new standalone languages.
This causes problems for potential users because building applications from components written in many different languages can be both unsafe and unnatural. 
An {internally-extensible} language could  address these issues, but designing a mechanism that is expressive while maintaining safety remains a challenge. 
Extensions must be modularly verified, their use in any combination must not weaken the metatheoretic properties of the language, nor can they interfere with one another.
We introduce a mechanism called active type-checking and translation (AT\&T) that aims to directly address these issues while remaining highly expressive. 
AT\&T leverages type-level computation, typed compilation techniques and a form of type abstraction to enable library-based implementations of a variety of primitives over a flexible  grammar in a safe manner. 
%In a language supporting AT\&T, primitive types families are represented uniformly at the type level. Each type is equipped with operators, defined using type-level functions, that are invoked specifically during the type-checking and translation of operations relevant to that type. 
%We discuss three points in the design space: (1) a simple judgmental construction of AT\&T designed to distill the essential concepts and admit formal safety theorems, (2) a "real" language called Ace that we use to demonstrate the expressive power of AT\&T in practice, and (3) a language design called Birdie that lifts and extends CoqÍs rich system of dependent types into the type-level, allowing extension developers to fully prove the correctness of extensions statically.
\keywords{extensible languages; active libraries; typed compilation; type-level computation; type abstraction}
\end{abstract}

\section{Motivation}
When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages}.
%\cite{delite}
Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules governing general-purpose abstractions (e.g. reasoning about units of measure requires built-in language support in F\# \cite{conf/cefp/Kennedy09}). 
Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract. For example, regular expressions encoded using inductive datatypes are often considered overly verbose, so most functional languages support them via strings, which is less safe. 
Finally, general-purpose abstractions are implemented in a uniform manner. Domain knowledge is not easily applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g. programmable GPUs) and distributed computing resources. Using variants of familiar abstractions backed by specialized implementation strategies can be quite useful in these cases.

@@

Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language of some form. They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}. 
%In our simple scenario, we may simply fork our implementation of Godel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). 
%In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs. Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}. Indeed, a number of software ecosystems are now explicitly designed to support many different languages, both general-purpose and domain-specific, atop a common intermediate language. The Java virtual machine (JVM), the Common Language Infrastructure (CLI) and LLVM are prominent examples of such ecosystems.

Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}. This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants. This compromises both verifiability and performance. For%requires the development of an interoperability layer for every pair of DSLs. 
Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change. This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction. We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a). 
%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
\begin{figure}[t]
\vspace{-15px}
\includegraphics[scale=0.415]{approaches.pdf}
\vspace{-30px}
\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g. JVM bytecode). Users can only interface with libraries written in another language via the constructs in the common language, causing \emph{interoperability problems}. (b) With the language-internal approach, the semantics of new abstractions (i.e. the logic that governs typechecking and translation to a fixed {internal language}, here labeled \texttt{I}) can be implemented directly within so-called \emph{active libraries}. Clients can import and use these abstractions directly whenever needed.}
\vspace{-10px}
\label{approaches}
\end{figure}

@@

\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries.
%Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code. 
As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thus achieve full safety, ease-of-use and performance due to the absence of wrappers and glue code. Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate. Libraries containing logic that is invoked at compile-time, as extension logic would be, have been called \emph{active libraries} \cite{activelibraries}. We adopt this terminology and diagram this competing approach in Figure \ref{approaches}(b).

@@

In this paper, we introduce a language-internal extensibility mechanism called \emph{active typechecking and translation} (AT\&T) that allows developers to introduce and implement the logic governing new primitive type families and operators from within libraries. 
We argue that this can be accomplished by enriching the type-level language, rather than introducing a separate metalanguage into the system. 
To make this proposal concrete, we begin by introducing a simple core calculus, called \atlam~(for the ``actively-typed lambda calculus''), in Section \ref{atlam}. 
This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and composability of extensions, as we outline in Section \ref{safety}.
In Section \ref{examples}, we suggest that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support.

@@

A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that do not require this level of control. To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic, without allowing so much control that the reliability of the system suffers. When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to the appropriate user-defined logic. Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification.

Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and type-level constructs like types, type aliases and datatype declarations. The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented. In fact, we will show that \textbf{the most natural place for type system extensions is within the type-level language}. The intuition is that extensions to a language's static semantics will need to manipulate types as values at compile-time. Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time (see Sec. \ref{related-work} for examples). The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation. This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two.

@@

												\\
\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
											    \kstr \pipe
												\kunit \pipe 
												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
												\kTypeBlur \pipe \kDen \pipe 
												\kIType \pipe \kITerm
												\\
\\												
%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%											 							&		&		&	\\
\textbf{internal terms} 				& 	\gamma	&	::=	&	\evar{x} \pipe 
												\ilam{\evar{x}}{\sigma}{\gamma} \pipe 
												\iapp{\gamma_{1}}{\gamma_{2}} \pipe
												\ifix{\evar{f}}{\sigma}{\gamma} \pipe
												\ipair{\gamma_{1}}{\gamma_{2}} \pipe 
												\ifst{\gamma} \pipe
												\isnd{\gamma}  
												\\
							& 		& 	\pipe	& 
												\iintlit \pipe \iop{\gamma_{1}}{\gamma_{2}} \pipe \iIfEq{\gamma_{1}}{\gamma_{2}}{\dint}{\gamma_{3}}{\gamma_{4}}  
												\\
 &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%\text{deabstracted}& \gamma & ::= & \mathcal{G}[\gamma, \sigma]\\
\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
												\dint \pipe
												\dpair{\sigma_1}{\sigma_2} \pipe
												\trepof{\tau} \pipe \dup{\tau}\\
\end{array}$$
\vspace{-10pt}
\caption{\small Syntax of \atlam. Variables $x$ are used in expressions and internal terms and are distinct from type-level variables, $\tvar{t}$. Names $\fvar{Fam}$ are family names (we assume that unique family names can be generated by some external mechanism) and $\opvar{op}$ are operator names. $\tstr{str}$ denotes string literals, $\iintlit$ denotes integer literals and $\oplus$ stands for binary operations over integers. %The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
\label{grammar}}
\end{figure}
In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise. 
\subsection{Overview}
The grammar of \atlam~is shown in Figure \ref{grammar}. 
A program, $\rho$, consists of a series of declarations followed by an expression. Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a declaration of a new primitive type family using \textsf{family}. Expressions, $e$, can be either variables, lambdas, or applications of operators, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$. This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.

The language is structured as a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
Types are type-level values of kind $\star$ (following System $F_{\omega}$) and classify expressions. The type-level language also includes other kinds of terms: type-level functions, lists (required by our mechanism), integers, strings and products for the sake of our examples (see Sec. \ref{termination} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.  The grammar of internal terms, $\gamma$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.

@@

\caption{G\"odel's T in \atlam, used to calculate 2+2. The statics we implement here are shown in the Appendix. Simple helper functions for working with lists ($\tvar{empty}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$),  described below, are given in the appendix.}

@@

To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}. These will be implemented internally as integers (that is, internal terms of internal type $\dint$). Figure \ref{example} shows how to define the indexed type family $\fvar{Nat}$. This family contains only one type, written $\ttype{Nat}{\tunit}$, which we alias on line \ref{nattype} by defining the type-level variable $\tvar{nat}$. We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) on lines \ref{z}-\ref{lastop} and use these to define a $plus$ function on line \ref{add} and compute 2+2.

@@

% CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
\bibliographystyle{abbrv}
\bibliography{../research}
\appendix
\section{Appendix}
\verb|http://www.cs.cmu.edu/~comar/att-esop14-appendix.pdf|.
\end{document}

>>>>

\documentclass[9pt,preprint]{sigplanconf}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
%\usepackage{cite}
%\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...

@@

\begin{document}

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

\title{Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism}

\authorinfo{~}{~}{~}
%\authorinfo{Cyrus Omar\and Jonathan Aldrich}
 %          {School of Computer Science\\
  %          Carnegie Mellon University}
   %        {\{comar, jonathan.aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
Researchers and domain experts often propose new language primitives as extensions to the semantics of an existing language. 
But today's statically-typed languages are {monolithic}: they do not expose language-internal mechanisms for implementing the static and dynamic semantics of new primitive types and their associated operators directly, so these experts must instead create new standalone languages.
This causes problems for potential users because building applications from components written in many different languages can be both unsafe and unnatural. 
An {internally-extensible} language could  address these issues, but designing a mechanism that is expressive while maintaining safety remains a challenge. 
Extensions must be modularly verified, their use in any combination must not weaken the metatheoretic properties of the language, nor can they interfere with one another.
We introduce a mechanism called active type-checking and translation (AT\&T) that aims to directly address these issues while remaining highly expressive. 
AT\&T leverages type-level computation, typed compilation techniques and a form of type abstraction to enable library-based implementations of a variety of primitives over a flexible  grammar in a safe manner. 
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Motivation}
When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a number of useful \emph{embedded domain-specific languages} \cite{delite}.
Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules that govern general-purpose abstractions (e.g. reasoning about units of measure requires built-in language support in F\# \cite{conf/cefp/Kennedy09}). 
Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract. For example, regular expressions encoded using inductive datatypes are quite verbose, so most functional languages support them via strings, which is less safe. 
Finally, general-purpose abstractions are implemented in a uniform manner, meaning domain-specific heuristics cannot be applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, such as when targeting heterogeneous hardware platforms (e.g. programmable GPUs) and distributed computing resources.

@@

Researchers or domain experts who run into situations like these, where more direct control over a language's semantics and implementation are needed, have little choice today but to realize new abstractions by creating a new language in some way. They might develop a new standalone language from scratch, modify an implementation of an existing language, or use tools like compiler generators, DSL frameworks and language workbenches \cite{fowler2010domain}. 
%In our simple scenario, we may simply fork our implementation of Godel's T or even edit it directly (a pernicious technique for implementing a new language where the prior one is overwritten). 
%In a more complex scenario, we may instead employ a tool like a compiler generator or DSL framework \cite{fowler2010domain} that can generate a standalone implementation from declarative specifications of language constructs. Some of these tools allow you to package and reuse these specifications (with the important caveat that not all combinations of constructs are valid and free of conflicts, an important modularity issue that we will return to several times in this paper).
The increasing sophistication and ease-of-use of these tools have led to calls for a {\it language-oriented approach} to software development, where different components of an application are written in different specialized languages \cite{journals/stp/Ward94}. Indeed, a number of software ecosystems are now designed explicitly to support many different languages, both general-purpose and domain-specific, atop a common intermediate language. The Java virtual machine (JVM), the Common Language Infrastructure (CLI) and LLVM are prominent examples of such ecosystems.

Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}. This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants. This compromises both verifiability and performance. %Forrequires the development of an interoperability layer for every pair of DSLs. 
Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change. This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction. We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a). 
%As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
\begin{figure*}[t]
\vspace{-15px}
\includegraphics[scale=0.415]{approaches.pdf}
\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g. JVM bytecode). Users can only interface with libraries written in another language via the constructs in the common language, causing \emph{interoperability problems}. (b) With the language-internal approach, the semantics of new abstractions (i.e. the logic that governs typechecking and translation to a fixed {internal language}, here labeled \texttt{I}) can be implemented directly within so-called \emph{active libraries}. Clients can import and use these abstractions directly whenever needed.}
\label{approaches}
\end{figure*}

@@

\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries.
%Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code. 
As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thereby achieve full safety, ease-of-use and performance. Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate. Libraries containing logic that is invoked at compile-time, as extension logic would be, have been called \emph{active libraries} \cite{activelibraries}. We adopt this terminology and diagram this competing approach in Figure \ref{approaches}(b).

@@

In this paper, we introduce a language-internal extensibility mechanism called \emph{active typechecking and translation} (AT\&T) that allows developers to introduce and implement the logic governing new primitive type and operator families from within libraries. 
We argue that this can be accomplished by enriching the type-level language, rather than introducing a separate metalanguage into the system. 
To make this proposal concrete, we begin by introducing a simple core calculus, called \atlam~(for the ``actively-typed lambda calculus''), in Section \ref{atlam}. 
This calculus uses type-level computation of higher kind, along with techniques borrowed from the typed compilation literature and a form of type abstraction that ensures that the implementation details of an extension are not externally visible to guarantee the safety of the language, the decidability of typechecking and compilation and non-interference of extensions, as we outline in Section \ref{safety}.
In Section \ref{examples}, we suggest that despite these constraints, this mechanism is expressive enough to admit, within libraries, a number of general-purpose and domain-specific abstractions that normally require built-in language support.

@@

A more appropriate and useful place for extensions like this is directly within libraries, alongside abstractions that can be adequately implemented in terms of existing primitive abstractions. To enable this, the language must allow for the introduction new primitive type families, like \lstinline{Prod}, operators, like \lstinline{Pair}, \lstinline{PrL} and \lstinline{PrR}, and associated typechecking and translation logic. When encountering these new operators in expressions, the compiler must effectively  hand control over typechecking and translation to the appropriate user-defined logic. Because this mechanism is {language-internal}, all compilers must support it to satisfy the language specification.

Statically-typed languages typically make a distinction between \emph{expressions}, which describe run-time computations, and type-level constructs like types, type aliases and datatype declarations. The design described above suggests we may now need to add another layer to our language, an {extension language}, where extensions can be declared and implemented. In fact, we will show that \textbf{the most natural place for type system extensions is within the type-level language}. The intuition is that extensions to a statically-typed language's semantics will need to manipulate types as values at compile-time. Many languages already allow users to write type-level functions for various reasons, effectively supporting this notion of types as values at compile-time (see Sec. \ref{related-work} for examples). The type-level language is often constrained by its own type system (where the types of type-level values are called \emph{kinds} for clarity) that prevents type-level functions from causing problems during compilation. This is precisely the structure that a distinct extension layer would have, and so it is quite natural to unify the two, as we will show in this work.

@@

												\\
\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
											    \kstr \pipe
												\kunit \pipe 
												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
												\kTypeBlur \pipe \kDen \pipe 
												\kIType \pipe \kITerm
												\\
\\												
%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%											 							&		&		&	\\
\textbf{internal terms} 				& 	\gamma	&	::=	&	\evar{x} \pipe 
												\ilam{\evar{x}}{\sigma}{\gamma} \pipe 
												\iapp{\gamma_{1}}{\gamma_{2}} \pipe
												\ifix{\evar{f}}{\sigma}{\gamma} \pipe
												\ipair{\gamma_{1}}{\gamma_{2}} \pipe 
												\ifst{\gamma} \pipe
												\isnd{\gamma}  
												\\
							& 		& 	\pipe	& 
												\iintlit \pipe \iop{\gamma_{1}}{\gamma_{2}} \pipe \iIfEq{\gamma_{1}}{\gamma_{2}}{\dint}{\gamma_{3}}{\gamma_{4}}  
												\\
 &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%\text{deabstracted}& \gamma & ::= & \mathcal{G}[\gamma, \sigma]\\
\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
												\dint \pipe
												\dpair{\sigma_1}{\sigma_2} \pipe
												\trepof{\tau} \pipe \dup{\tau}\\
\end{array}$$
\vspace{-10pt}
\caption{\small Syntax of \atlam. Variables $x$ are used in expressions and internal terms and are distinct from type-level variables, $\tvar{t}$. Names $\fvar{Fam}$ are type family names (we assume that globally unique type family names can be generated by some external mechanism) and $\opvar{op}$ are operator family names. $\tstr{str}$ denotes string literals, $\iintlit$ denotes integer literals and $\oplus$ stands for binary operations over integers. %The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
\label{grammar}}
\end{figure}
In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise. 
\subsection{Overview}
The grammar of \atlam~is shown in Figure \ref{grammar}. 
A program, $\rho$, consists of a series of declarations followed by an expression. Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a primitive type family declared using \textsf{family}, which can contain implementations of one or more operator families, $\theta$. Expressions, $e$, can be either variables, lambdas, or applications of operators.
%, and are ultimately given meaning by translation to a typed internal language, with terms $\gamma$ and types $\sigma$. 
%This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language.

The language is structured as a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
Types are type-level values of kind $\star$ (following System $F_{\omega}$) and classify expressions. The type-level language also includes other kinds of terms: type-level functions, lists (required by our mechanism), integers, strings and products for the sake of our examples (see Sec. \ref{safety} for a discussion on other acceptable kinds of type-level data) and constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.

All expressions are given meaning by translation to a typed internal language. This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language. The grammar of internal terms, $\gamma$, and internal types, $\sigma$, also includes special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends.

@@

\caption{G\"odel's T in \atlam, used to calculate 2+2. Helper functions for working with lists ($\tvar{empty}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$),  described below, are given in the appendix. We will refer to this program as $\rho_{\text{nat}}$ in the text.}

@@

To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}. These will be implemented internally as integers (that is, internal terms of internal type $\dint$). Figure \ref{example} shows how to define the indexed type family $\fvar{Nat}$. This family contains only one type, written $\ttype{Nat}{\tunit}$, which we alias on line \ref{nattype} by defining the type-level variable $\tvar{nat}$. We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) on lines \ref{z}-\ref{lastop} and use these to define a $plus$ function on line \ref{add} and compute 2+2. We will refer back to this figure as we describe each construct below.

@@

% CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
\bibliographystyle{abbrv}
\bibliography{../research}

%\acks
%The author is grateful to Jonathan Aldrich, Robert Bocchino and anonymous referees for their useful suggestions. This work was funded by the DOE Computational Science Graduate Fellowship under grant number DE-FG02-97ER25308.
%Acknowledgments, if needed.

\end{document}
